---
title: "Final Project"
author: "Caroline Ledbetter"
header-includes:
- \usepackage{setspace}
- \doublespacing
- \usepackage{amsmath}
output:
  pdf_document:
    fig_caption: yes
    fig_height: 4
    fig_width: 5
fontsize: 12pt
geometry: right=1in, left=1in, top=1.25in, bottom = 1.25in
csl: /Users/Caroline/Repositories/causal_analytics/Docs/american-medical-association.csl
bibliography: /Users/Caroline/Repositories/causal_analytics/Docs/paper.bib
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE)
```


```{r setup, include=FALSE}
load(file = "~/Repositories/Data/Causal_Analytics/dataclean.rdata")
load(file = "~/Repositories/Data/Causal_Analytics/analysis.rdata")
library(pander)
```

# Study Background  
The data for this project comes from the Surveillance, Epidemiology and End Results (SEER) Program of the National Cancer Institute. SEER collects information on incidence, survival, and prevalence from specific geographic areas representing 26 percent of the US population. The SEER Program is the only comprehensive source of population-based information in the United States that includes stage of cancer at the time of diagnosis and patient survival data. [@SEER] This study specifically explores Lymphoma, a cancer that begins in the lymph system. The treatment and outcome of lymphoma are dependent on the stage and type of lymphoma. [@NCI]  This study aims to explore differences in survival of patients treated with radiation and surgery vs radiation only.    

# Methods and Results  
*Methods*
Data were collected from three SEER regions (1, 2, and 20). Information on birth year, age, sex, grade of lymphoma (one to nine, with higher grades indicating more aggressive lymphoma), site (Hodgkin or non-Hodgkin), race (white or black), and SEER stage (localized, regional, distant, or unknown) were collected. Individuals who reported they did not receive radiation therapy after cancer diagnosis were excluded. Follow up time in days, survival status at the end of follow-up and cause of death were also recorded.   Treatment was defined as radiation with surgery, controls were defined as radiation without surgery.

A propensity score for treatment was calculated using all available covariates. Propensity scores were calculated using logistic regression. An inverse probability weight was calculated for each subject, $w_i =  \frac{Z_i}{e_i}  +  \frac{1-Z_i}{1-e_i}$, where *Z~i~* = an indicator variable denoting whether the or not the *i*th subject was treated and *e~i~* is the propensity score.[@austin]  The inverse probability weight was then used to obtain balanced treatment and control groups. Table 1 shows the participant characteristics by treatment group with and with out weighting. To determine if covariates varied across treatment groups, categorical variables were tested using $\chi ^2$ and continuous variables were tested using ANOVA. For the weighted analysis, continuous variables were tested using logistic regression. Inverse probability of treatment weighting using the propensity score allows us to obtain an unbiased marginal treatment effect or an estimate of the average treatment effect using cox hazard regression [@austin].

Cox hazard regression was used to obtain a hazard ratio for all-cause mortality and death from lymphoma. Weights were incorporated using the survey package v(`r packageVersion('lmerTest')`) [@survey] in `r R.Version()$version.string` [@R]. A sensitivity analysis was performed.   

*Results*
The data contained `r format(no_rad + nrow(lymph), big.mark = ',', trim = T)` observations, `r format(no_rad, big.mark = ',', trim = T)` were excluded because they did not receive radiation treatment, leaving `r format(nrow(lymph), big.mark = ',', trim = T)` for analysis. Participants receiving treatment were more likely to have Hodgkin lymphoma and to be from location 20, treated participants were also more likely to be black and be younger. Inverse probability weighting resulted in balanced treatment groups. (Table 1). The maximum follow-up time was `r max(lymph$Tim)` days. Individuals who received treatment were more likely to survive to the end of follow-up (Fig 1). The estimated hazard ratio for all cause mortality was `r format(exp(model$coefficients), digits = 2, nsmall = 2)` (95% CI: `r format(exp(model$coefficients - 1.96*sqrt(model$var)), digits = 2, nsmall =2)` - `r format(exp(model$coefficients + 1.96*sqrt(model$var)), digits = 2, nsmall =2)`), thus undergoing surgery in addition to radiation reduced the hazard of all cause mortality by `r format((1 - exp(model$coefficients))*100, digits = 0)`%. The estimated hazard ratio for death due to lymphoma was `r format(exp(model_lymdth$coefficients), digits = 2, nsmall = 2)` (95% CI: `r format(exp(model_lymdth$coefficients - 1.96*sqrt(model_lymdth$var)), digits = 2, nsmall =2)` - `r format(exp(model_lymdth$coefficients + 1.96*sqrt(model_lymdth$var)), digits = 2, nsmall =2)`), thus undergoing surgery in addition to radiation reduced the hazard of death due to lymphoma by `r format((1 - exp(model_lymdth$coefficients))*100, digits = 0)`%.  

*Discussion* Adjusting for covariates in the model, as is typically done in an observational study estimates a conditional treatment effect, or the average effect at an individual level of changing a subject's treatment status. The marginal treatment effect is the difference in outcomes between a population where everyone is treated and the counter-factual where no one is treated. Propensity score methods estimate marginal effects. This allows us estimate causal treatment effects in observational studies. [@austin] These causal effects estimates however are based on the assumption that there is no unmeasured confounding. Unmeasured covariates that effect the treatment assignment and the outcome can cause us to make inaccurate estimates. A sensitivity analysis where we evaluate different levels of an unmeasured confounding on our results can help us determine the extent to which this is a concern. In a randomized control trial, because treatment assignment is random, unmeasured variables that effect the outcome are unlikely to not be balanced across treatment groups. In an observational study, because we may not be aware of the functional form of treatment assignment, this is of greater concern. An observational study with a large sample size is likely to have high precision and thus low p-values. However, the opportunity for high bias exists. We used propensity score modeling to minimize this bias.  

# Conclusions  
Undergoing surgery in addition to radiation resulted in significantly improved outcomes and decreased the risk of all-cause mortality and lymphoma cause death. Participants who received surgery in a sequence with radiation had a `r format((1 - exp(model$coefficients))*100, digits = 0)`% reduction in all cause mortality when compared to those who received radiation [HR: `r format(exp(model$coefficients), digits = 2, nsmall = 2)` (95% CI: `r format(exp(model$coefficients - 1.96*sqrt(model$var)), digits = 2, nsmall =2)` - `r format(exp(model$coefficients + 1.96*sqrt(model$var)), digits = 2, nsmall =2)`)]. Participants who received surgery in a sequence with radiation had a `r format((1 - exp(model_lymdth$coefficients))*100, digits = 0)`% reduction in lymphoma caused death when compared to those who received radiation [HR: `r format(exp(model_lymdth$coefficients), digits = 2, nsmall = 2)` (95% CI: `r format(exp(model_lymdth$coefficients - 1.96*sqrt(model_lymdth$var)), digits = 2, nsmall =2)` - `r format(exp(model_lymdth$coefficients + 1.96*sqrt(model_lymdth$var)), digits = 2, nsmall =2)`)]. 

# References   

